---
title: "Class 15: Transcriptomics and the analysis of RNA-Seq data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import countData and colData into R
```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
```

```{r}
dim(counts)
head(counts)
```

```{r}
head(metadata)
```

```{r}
colnames(counts)

```

```{r}
metadata$id
```

#Check for correspondence of metadata and countdata
```{r}
#[-1]: everything but the first
all(colnames(counts)[-1] == metadata$id)
#if any are false -> false
```

First tast: find which columns in **counts** correspond to the control condition (i.e. no drug!) Then we can calculate the mean count value for each gene.

```{r}
control.inds <- metadata$dex == "control"
control.ids <- metadata$id[control.inds]
head(counts[,control.ids])
```

Calculate the mean value for all genes
```{r}
control.mean <- (rowSums(counts[,control.ids])/ncol(counts[,control.ids]))
names(control.mean) <- counts$ensgene
head(control.mean)
```

Calculate the mean for the drug treated


```{r}
#drug.inds <- metadata$dex == "treated"
#drug.ids <- metadata$id[drug.inds]
treated <- metadata[metadata[,"dex"]=="treated",]
treated.mean <- rowSums(counts[,treated$id])/ncol(counts[,treated$id])
#treated.mean <- (rowSums(counts[,drug.ids])/ncol(counts[,drug.ids]))
names(treated.mean) <- counts$ensgene
head(treated.mean) 
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

```{r}
colSums(meancounts)
```


Now we can plot the drug treated vs non-drug control count values

```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

```{r}
plot.default(meancounts$control.mean, meancounts$treated.mean, log = "xy")
```


We can find candidate differentially expressed genes by looking for genes with a large change between control and dex-treated samples. We usually look at the log2 of the fold change, because this has better mathematical properties.

Here we calculate log2foldchange, add it to our meancounts data.frame and inspect the results either with the head() or the View() function for example.
```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```
There are a couple of “weird” results. Namely, the NaN (“not a number””) and -Inf (negative infinity) results.

The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. It turns out that there are a lot of genes with zero expression. Let’s filter our data to remove these genes. Again inspect your result (and the intermediate steps) to see if things make sense to you

```{r}
toy <- c(1,2,0,4,0)
which(toy == 0)
```

```{r}
toy <- meancounts[1:6,]
toy[,1:2]
which(toy[,1:2] == 0, arr.ind = TRUE)
unique(which(toy[,1:2] == 0, arr.ind=TRUE)[,1])
```


Remove the zero entry genes from meancounts to produce **mycounts**
```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```
A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.


```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

```{r}
sum(up.ind)
```

```{r}
sum(down.ind)
```

The are 'sum(up.ind)' number of up regulated genes and 'sum(down.ind)' down regulated genes.

```{r}
head(mycounts[up.ind,])
```

##adding annotation data

use the **merge()** function to add annotation data from ensemble

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

```{r}
mycounts.annotated <- merge(x=mycounts, y=anno, by.x="row.names", by.y="ensgene")
head(mycounts.annotated)
```

##DeSeq2 analysis

Now we do this analysis properly with p-values!

```{r}
library(DESeq2)

```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```


```{r}
res <- results(dds)
res
```


```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
res01 <- results(dds, alpha=0.01)

```


##make a volcano plot
```{r}
mycols <- rep("gray", nrow(res05))
mycols[abs(res05$log2FoldChange)>2] <- "blue"
plot(res05$log2FoldChange, -log(res$padj), col = mycols)
```

```{r}
mycols <- rep("gray", nrow(res01))
mycols[abs(res01$log2FoldChange)>2] <- "red"

inds <- (res01$padj < 0.01) & (abs(res01$log2FoldChange) > 2)
mycols[inds] <- "blue"

plot(res01$log2FoldChange, -log(res01$padj), col = mycols, ylab = "-Log(P-value)", xlab = "Log2(FoldChange)")

abline(v = c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col ="gray", lty=2)
```


