---
title: "Class11: Structural Bioinformatics 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1. The PDB

Q1: Determine the percentage of structures solved by X-Ray and Electron Microscopy. From
the website what proportion of structures are protein? Aim to have a rendered GitHub document
with working code that yields your answers.
```{r}
stats <- read.csv("Data Export Summary.csv", row.names = 1)
stats
```

```{r}

percent.by.method <- stats$Total/sum(stats$Total)*100
names(percent.by.method) <- rownames(stats)
percent.by.method

```

```{r}
sum(stats$Proteins)/sum(stats$Total)*100

```


## Section 3. Introduction to Bio3D in R

First we have to load the package we want to use
```{r}
library(bio3d)
```

Now we can use the functions from this bio3d package

```{r}
pdb <- read.pdb("1hsg")
pdb
```

```{r}
#head(aa321(pdb$atom$resid))

```



```{r}
ca.inds <- atom.select(pdb, "calpha")
ca.inds
```

```{r}
# Print details of the first few selected atoms
head( pdb$atom[ca.inds$atom, ] )

# And selected xyz coordinates
head( pdb$xyz[, ca.inds$xyz] )
## [1] 64.044 51.615 33.423 62.439 54.794 32.359
```


In addition to the common selection strings (such as ‘calpha’ ‘cbeta’ ‘backbone’ ‘protein’
‘notprotein’ ‘ligand’ ‘water’ ‘notwater’ ‘h’ and ‘noh’) various individual atom properties can be
used for selection.
```{r}
# Select chain A
a.inds <- atom.select(pdb, chain="A")
# Select C-alphas of chain A
ca.inds <- atom.select(pdb, "calpha", chain="A")
# We can combine multiple selection criteria to return their intersection
cab.inds <- atom.select(pdb, elety=c("CA","CB"), chain="A", resno=10:20)
cab.inds

```

We want to select out the protein and drug only parts of these molecular files
```{r}
prot.inds <- atom.select(pdb, "protein")
prot.inds

```

```{r}
prot.pdb <- trim.pdb(pdb, prot.inds)
write.pdb(prot.pdb, file = "protein.pdb")
```

Let's do the same things for our drug
```{r}
lig.inds <- atom.select(pdb, "ligand")
lig.pdb <- trim.pdb(pdb, lig.inds)
write.pdb(lig.pdb, file = "ligand.pdb")

```

