---
title: "Class 8 Clustering and PCA"
author: "Yvette Tan"
date: "2/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means Clustering

Use the kmeans() function setting k to 2 and nstart=20
Inspect/print the results
Q. How many points are in each cluster?

Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points
```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

```{r}
km <- kmeans(x, centers = 2, nstart = 20)
```

```{r}
km
```

Q. What ‘component’ of your result object details

Cluster size
```{r}
km$size
```

Cluster membership and assignment
```{r}
km$cluster
```

Cluster center
```{r}
plot(x, col = km$cluster)
points(km$centers, col = "blue", pch = 15, cex =2)
```

```{r}
km$totss
```


## Hierarchical Clustering

```{r}
# First we need to calculate point (dis)similarity
# as the Euclidean distance between observations
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
# clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
hc

```

```{r}
plot(hc)
abline(h = 6, col = "red")
cutree(hc, h=6)
```

We can also use k=number of cluster groups I want with **cutree()**

```{r}
cutree(hc, k=2)
```

## Different Linkage Methods


```{r}
#Using different hierarchical clustering methods
d <- dist_matrix
hc.complete <- hclust(d, method="complete")
plot(hc.complete)
hc.average <- hclust(d, method="average")
plot(hc.average)
hc.single <- hclust(d, method="single")
plot(hc.single)
```


Run hclust() on this made up overlapping data
```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```


Q. Use the dist(), hclust(), plot() and cutree()
 functions to return 2 and 3 clusters
 
```{r}
dist_matrix <- dist(x)
# The hclust() function returns a hierarchical
# clustering model
hc <- hclust(d = dist_matrix)
# the print method is not so useful here
plot(hc)
abline(h=2, col = "red")
abline(h=2.5, col = "blue")

```

```{r}
gp2 <- cutree(hc, k=2)
gp3 <- cutree(hc, k=3)
```

```{r}
plot(x, col = gp3)
```


```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV",
 row.names=1)
head(mydata)
```

```{r}
#head(t(mydata))
pca <- prcomp(t(mydata), scale=TRUE)
pca
```

```{r}
plot(pca$x[,1], pca$x[,2])
```


```{r}
##precent varience is often more informative to look at
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
## Click to identify which sample is which
identify(pca$x[,1], pca$x[,2], labels=colnames(mydata))
## Press ESC to exit… 
```



## NEW
```{r}
y <- read.csv("UK_foods.csv",
 row.names=1)
y
```
## Complete the following code to find out how many rows and columns are in x?

```{r}
dim(y)
#or ncol(x) or nrow(x)
```

```{r}
head(y)
```


```{r}
barplot(as.matrix(y), beside=T, col=rainbow(nrow(y)))
```
Q3: Changing what optional argument in the above barplot() function results in the following plot?
changing beside = FALSE

```{r}
barplot(as.matrix(y), beside=FALSE, col=rainbow(nrow(y)))
```

Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?
```{r}
pairs(y, col=rainbow(10), pch=16)

```

##PCA

```{r}
pca <- prcomp( t(y) )
summary(pca)
```

plot of pc1 vs pc2

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(y), col = c("orange", "red", "blue", "darkgreen"))
```


```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
z <- summary(pca)
z$importance
```




##Examine the Loadings

This will hep us determine how the original variables (dimensions) contribute to our new PCs
```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```


for PC2?
```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2)
```





