---
title: "Class 13: Structure based drug discovery"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prepare HIV-protease molecule for docking

First get a PDB structure from the online database then clean it up for docking.
```{r}
library(bio3d)
file.name <- get.pdb("1hsg")
```

Read file into R and clean it up (remove non-protein bits) and also we will save the ligand into a separate file
 

```{r}
hiv <- read.pdb(file.name)
hiv

```

Q1: What is the name of the two non protein resid values in this structure? What does resid
correspond to and how would you get a listing of all reside values in this structure?

the two non protein resid values: H2O and MK1

resid: residues

```{r}
prot <- trim.pdb(hiv, "protein")
prot
```

```{r}
write.pdb(prot, file = "1hsg_protein.pdb")
```


```{r}
lig <- trim.pdb(hiv, "ligand")
write.pdb(lig, file = "1hsg_ligand.pdb")
```

Docking

In order to visualize the docks and compare to the crystal conformation of the ligand we will
process the all.pdbqt to a PDB format file that can be loaded into VMD. To do this we will use R
and the Bio3D package.

```{r}
library(bio3d)
res <- read.pdb("all.pdbqt", multi=TRUE)
write.pdb(res, "results.pdb")
```

##Normal Mode Analysis

```{r}
pdb <- read.pdb("1HEL")
#normal mode analysis
m <- nma(pdb)
plot(m)
```


```{r}
mktrj(m, mode = 7, file = "nma_7.pdb")
```

